---
title: "Getting started with DrWrinch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with DrWrinch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(DrWrinch)
```

## The problem

A researcher classifying within-case evidence for a working theory
($H_1$) against a single rival ($H_R$) wants to summarize the weight of
evidence with a Bayes factor. The challenge is that the per-observation
probabilities a Bayes factor requires are hard to motivate without
overcommitting. DrWrinch supplies two fully specified generative models
that produce those probabilities, both tilted against $H_1$ by
construction, so the reported Bayes factor is a lower bound on the
evidence in favor of the working theory.

This vignette works the running example from Lopez, Bowers, and Gajardo
Cooper (2026): nine pieces of evidence favoring $H_1$ and three
favoring $H_R$.

## Two models

**Binomial model.** Each observation is an independent Bernoulli draw
from an infinite evidence universe with success probability $\theta$.
$H_1$ corresponds to $\theta > 1/2$. Under a uniform prior on $\theta$,
the Bayes factor is the ratio of posterior masses on the two sides of
$1/2$. This model fits research designs where the evidence universe is
open-ended: ongoing interviews, an expanding archive.

```{r}
bf_binomial(y_W = 9, y_R = 3)
```

**Hypergeometric urn model (Formulation C).** Evidence is drawn without
replacement from a finite urn. The Working Theory Favorable (WTF)
sub-model has urn composition $(y_W + 1, \max(1, y_R))$; the Rival
Theory Favorable (RTF) sub-model has urn composition $(y_W, y_W + 1)$.
The construction adds one unobserved pro-$H_1$ item to the working
theory's urn and tilts the rival's urn toward $H_R$ by exactly one
item. This model fits research designs where the evidence base is
bounded: a closed historical archive, a fixed roster of documents.

```{r}
bf_urn(y_W = 9, y_R = 3)
```

The hypergeometric BF of 323 has a closed form:
$(10/13) / (120/50388) = 323$.

## When does each model apply?

| Design | Model |
|---|---|
| Ongoing interviews, growing archive, open-ended fieldwork | `bf_binomial()` |
| Closed historical archive, fixed roster of documents | `bf_urn()` |

When in doubt, compute both. They report different but compatible
quantities: the binomial conditions on the evidence universe being
infinite; the urn conditions on it being bounded.

## A property worth knowing

Under symmetric evidence ($y_W = y_R$), both Bayes factors are exactly
$1$ -- no evidence in either direction.

```{r}
bf_binomial(5, 5)
bf_urn(5, 5)
```

For `bf_urn()` this is the central property distinguishing Formulation
C from naive constructions: the urn model returns 1 rather than $\infty$
or $0$ at the equipoise point.

## Sensitivity to observation bias

A reviewer might ask: how robust is the conclusion to the possibility
that pro-$H_1$ items were more likely to be observed than pro-$H_R$
items? `sens_urn()` reports the smallest observation-bias factor
$\omega > 1$ at which the Bayes factor first drops below a decision
threshold (the default threshold is $20$, the conventional cutoff for
"strong" evidence).

```{r}
s_urn <- sens_urn(9, 3, threshold = 20)
s_urn$bf
s_urn$omega_star
```

The reading: pro-$H_1$ items would have had to be roughly 2.4 times
more likely to be noticed than pro-$H_R$ items before the urn
conclusion reverses.

## Sensitivity to coding error

A peer who cannot re-read every document can still ask how many of the
pro-$H_1$ observations would have to be re-coded as pro-$H_R$ before the
conclusion changes. `sens_coding()` reports that tipping point for either
model: re-coding $x$ observations moves the counts from $(y_W, y_R)$ to
$(y_W - x, y_R + x)$ and recomputes the same Bayes factor.

```{r}
sens_coding(9, 3, model = "binomial", threshold = 20)$x_star
sens_coding(9, 3, model = "urn", threshold = 20)$x_star
```

One re-coding overturns the binomial conclusion and two overturn the
hypergeometric. The binomial tolerates less because its Bayes factor
(20.67) starts only just above the threshold; the urn starts at 323 and
so absorbs one more re-coding before crossing.

## Sensitivity to the prior on $\theta$

For the binomial model, `sens_binomial()` additionally reports the
smallest rival-tilted prior at which the conclusion reverses. The prior
$\text{Beta}(1, M + 1)$ posits $M$ pseudo-observations all favoring the
rival; `M_star` is the smallest such $M$ that drives the Bayes factor
below threshold.

```{r}
s_binom <- sens_binomial(9, 3, threshold = 20)
s_binom$bf
s_binom$omega_star
s_binom$M_star
```

At the running example the binomial Bayes factor only just clears the
threshold, so `M_star` is 1: a single rival-favoring pseudo-observation
in the prior drops it below 20. A larger evidence base would tolerate
more.

## Weighted evidence

A researcher who believes one observation carries more probative weight
than the others can sum integer weights and pass the totals:

```{r}
# Eight ordinary pro-H1 items at weight 1 and one "smoking gun" at
# weight 10, matching the paper's weighted running example: W = 8 + 10 = 18.
w_W <- c(10, rep(1, 8))
w_R <- rep(1, 3)
bf_binomial(sum(w_W), sum(w_R))
bf_urn(sum(w_W), sum(w_R))
```

The "+1" rival-favoring construction goes through unchanged. See the
paper's discussion of probative weight for the rationale.
