---
title: "Getting started with DrWrinch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with DrWrinch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(DrWrinch)
```

## The problem

A researcher has read her sources and coded each piece of evidence as
supporting her working theory or supporting a single rival explanation.
She has nine of the first kind and three of the second. She wants to
report how much those twelve observations favor her theory over the
rival, and she wants a critic to be unable to say she picked the
arithmetic to suit herself.

A Bayes factor is the ratio of how probable her counts are under one
theory to how probable they are under the other. Getting that ratio
takes a probability for the counts under each theory, and a theory
stated in words does not supply one. This package supplies them from a
single model, and this vignette works the running example of Lopez,
Bowers, and Gajardo Cooper (2026): nine observations supporting the
working theory, three supporting the rival.

## One model

Suppose each observation she might have read supports the working
theory with probability $\theta$. Read $\theta$ as the share of the
available evidence that supports her theory. Her working theory claims
that share is above one half. The rival claims it is at or below one
half.

That is the whole model. It says nothing about how much evidence exists,
which matters: she never has to state the size of an archive she has not
counted.

Under it, the probability of seeing $k$ supporting observations out of
$N$ at a given share is the binomial probability, so the numbers below
are things a reader can recompute.

## Why one model gives two Bayes factors

Her working theory names a range of shares, and so does the rival. A
range is not a probability for the counts. To get one, she has to say
how she spreads her weight over the shares inside each range, and the
package's default gives every share in a range the same weight.

For her own theory that settles it: average the probability of her
counts over the shares above one half. That average is the numerator of
both Bayes factors below.

For the rival there are two answers, and neither can be called
arbitrary or self-serving. She can hold the rival to every version of
his claim equally, averaging over his whole range the same way she
averaged over hers. Or she can hand him the single share in his range
that makes her counts most probable, which is one half, and see whether
her evidence survives even that generosity.

Both denominators, and the numerator they share, are three numbers she
can compute:

```{r}
n <- 12
k <- 9

# Averaging over a range half as wide as the interval from 0 to 1 means
# dividing the integral by one half, which is the factor of 2 below.
numerator <- 2 * integrate(function(t) dbinom(k, n, t), 0.5, 1)$value
den_avg   <- 2 * integrate(function(t) dbinom(k, n, t), 0, 0.5)$value
den_half  <- dbinom(k, n, 0.5)

round(c(numerator = numerator, den_avg = den_avg, den_half = den_half), 5)
```

Dividing the shared numerator by each denominator gives the two Bayes
factors, and the package computes each directly:

```{r}
c(by_hand = numerator / den_avg, from_package = bf_uniform_weights(9, 3))
c(by_hand = numerator / den_half, from_package = bf_worst_case(9, 3))
```

So her twelve observations give 20.67 when the rival is held to his
whole range and 2.73 when he is handed his best single share. The two
differ by a factor of 7.57, and because the numerator cancels, that
factor is just the ratio of the two denominators:

```{r}
round(c(ratio_of_bayes_factors = (numerator / den_avg) / (numerator / den_half),
        ratio_of_denominators = den_half / den_avg), 2)
```

Reading that aloud: nine of twelve is 7.57 times as probable at a share
of exactly one half as it is on average across the shares at or below
one half. Averaging over the rival's whole range includes shares near
zero, under which nine of twelve is very improbable indeed, and it is
those shares that lift 2.73 to 20.67.

## What the smaller number buys

The generous reading is worth more than its size suggests, for two
reasons.

It is a lower bound. One half is the share in the rival's range that
fits her counts best, so no weighting of that range she could have
written down before coding would make the evidence against him look
weaker. A critic who says "you chose weights that flattered you" has
nothing to point at.

It also comes with an error rate. In a world where the rival is right,
the average of this Bayes factor across repeated samples is exactly one:

```{r}
lambda <- sapply(0:12, function(k) bf_worst_case(k, 12 - k))
sum(dbinom(0:12, 12, 0.5) * lambda)
```

A quantity that is never negative and averages one cannot exceed 20 more
than one time in twenty. So a researcher who treats 20 as the value at
which she sets the rival aside is misled at most one time in twenty when
he is right. The guarantee holds for evidence drawn independently and
for evidence drawn without replacement from a body of any size, so she
still never has to say how large that body is.

The price is that 20 is demanding. Out of twelve observations, the
smallest count that reaches it is eleven:

```{r}
min(which(lambda >= 20)) - 1
```

## When the counts do not reach the threshold

At nine of twelve the worst-case Bayes factor is 2.73, so the researcher
never sets the rival aside at a threshold of 20 on that reading. Rather
than stop there, she can report how far apart the two theories' claims
would have to be before her counts do reach 20. `separation_g()` solves
for that distance:

```{r}
g <- separation_g(9, 3, threshold = 20)
c(separation = g, working_theory_claims = 0.5 + g, rival_claims = 0.5 - g)
```

Her twelve observations reach 20 once her theory claims at least 62.3
percent of the evidence and the rival concedes at most 37.7 percent. She
did not choose that distance; she solved for it. `bf_separated()` shows
that the distance does what it says:

```{r}
bf_separated(9, 3, g)
```

## Choosing weights other than the uniform

Giving every share in a range the same weight is a choice, and a reader
may want a different one. `bf_rescaled()` takes a Beta distribution
squeezed onto each theory's half, one pair of parameters per side, and
the uniform default is the pair $(1, 1)$ on both sides.

The parameters say where the weight piles up. Below one, weight
collects at both ends of each half, including right next to one half
where the two theories claim nearly the same share; above one, it
collects in the middle of each half, so the two theories claim shares
far apart.

```{r}
c(uniform      = bf_rescaled(9, 3),
  next_to_half = bf_rescaled(9, 3, 0.5, 0.5, 0.5, 0.5),
  far_apart    = bf_rescaled(9, 3, 10, 10, 10, 10))
```

Weights that place the two theories nearly on top of each other leave
her counts less to distinguish, and the value falls from 20.67 to 9.53.
Weights that push the theories apart, so hers predicts a share near
three quarters and the rival's near one quarter, raise it to 252.92. The
weights are a substantive claim about what each theory predicts, and a
reader is entitled to argue with them.

## What a reader would have to grant

Three questions a critic can press, each answered by the value at which
her report would change.

**How she coded the evidence.** A peer cannot re-read every document,
but he can ask how many of the nine supporting observations would have
to be re-coded as supporting the rival:

```{r}
sens_coding(9, 3, threshold = 20)$x_star
```

One. The uniform-weights Bayes factor tolerates so little because 20.67
sits barely above 20 to begin with.

For the worst-case Bayes factor the question changes, because re-coding
cannot overturn a conclusion she did not draw. What re-coding moves is
the distance she reports, and each re-coding narrows the margin between
the counts by two:

```{r}
sens_coding(9, 3, model = "worst_case", threshold = 20)$recoding
```

After one re-coding her theory would have to claim 67.9 percent of the
evidence rather than 62.3, and after three the counts are even and no
distance reaches 20 at all.

**How evenly she searched.** If her search was more likely to surface
evidence for her own theory, part of the counts' dominance came from how
she looked:

```{r}
s <- sens_binomial(9, 3, threshold = 20)
s$omega_star
```

Supporting evidence would have to be about one percent likelier to
surface than opposing evidence before the value falls to 20.

**How much weight she gave the rival before starting.**
`sens_binomial()` also sweeps a Beta$(1, M + 1)$ prior on the whole
interval from zero to one, which reads as $M$ background cases all
favoring the rival:

```{r}
s$M_star
```

One background case is enough. Read that number carefully: what falls as
$M$ grows is the posterior odds, not the Bayes factor. A prior on the
whole interval says how much weight each theory gets as well as how
weight is spread inside each theory's range. The Bayes factor cancels
the first away and the posterior odds keep it, so the two separate as
soon as the prior stops being uniform, and they move in opposite
directions:

```{r}
three_objects <- function(M) {
  lik <- function(t) dbinom(9, 12, t)
  num <- integrate(function(t) lik(t) * dbeta(t, 1, M + 1), 0.5, 1)$value
  den <- integrate(function(t) lik(t) * dbeta(t, 1, M + 1), 0, 0.5)$value
  pi_1 <- 1 - pbeta(0.5, 1, M + 1)
  c(M = M,
    prior_odds = pi_1 / (1 - pi_1),
    bayes_factor = (num / pi_1) / (den / (1 - pi_1)),
    posterior_odds = num / den)
}
round(t(sapply(0:4, three_objects)), 2)
```

Tilting toward the rival commits him to shares near zero, under which
nine of twelve is more surprising still, so the Bayes factor rises. The
prior odds fall faster, so the posterior odds fall. `bf_binomial()`
returns that last column, which is why `M_star` is a tipping point in
the posterior odds.

## Observations that carry more weight than others

A researcher who judges one document more probative than the rest can
enter it as several observations by summing integer weights:

```{r}
# Eight ordinary supporting observations plus one she weights at ten.
w_W <- c(10, rep(1, 8))
w_R <- rep(1, 3)
c(uniform_weights = bf_uniform_weights(sum(w_W), sum(w_R)),
  worst_case      = bf_worst_case(sum(w_W), sum(w_R)))
```

These are weights on observations, which act as replication, and they
are a different thing from the weights on shares that `bf_rescaled()`
takes. See the paper's discussion of probative weight for when a
researcher can defend one.
